Reproducible development environments with rix

### 2025-02-19 paper.qmd
continue on 630

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of R needs to be handled
separately. [renv]{.pkg} will also not install system-level dependencies such
as [cmake]{.pkg} for [nloptr]{.pkg} on Linux or macOS, so these still need to
be handled separately, as described in [renv]{.pkg}’s Readme: 
\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

@@

The `.Rprofile` file gets read by R when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror. However, unless the user explicitly and
manually manages the different projects' libraries, using the Posit Package
Manager will install all the package for all the projects at the given date.

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.pkg} by the R infrastructure team [-@rlib2023].
[rig]{.pkg} makes it easy to install different versions of [R]{.proglang},
so one could use it to install the right version to reproduce an analysis, and
then use one of the listed packages above to install the right library of
[R]{.proglang} packages. However, this involves many manual steps and is thus
error prone and time-consuming.

@@

A common approach seems to be to work on a study interactively using a standard
installation of R and R packages, and after the study is done, take some time to
write the required `Dockerfile` to provide an easy way for future replicators to
reproduce the results of the study. But this approach does not solve the problem
of deploying the same environment on several machines, and so ideally authors
would develop using a containerized development environment. Another challenge
of using [Docker]{.pkg} is that, as stated already, [Docker]{.pkg} images are
actually almost complete Linux operating systems, so experience with Linux is
highly recommended if one wishes to properly write a `Dockerfile`. Finally, one
also needs to be careful and write the `Dockerfile` in such a way that the build
process will always result in the same image. This is only possible if a
specific digest of a base image is used instead of a tagged version, but in
practice, this is rarely done.

@@

## The Nix package manager {#sec-nix}

[Nix]{.pkg} is a package manager that can be used to install and build software
in a completely reproducible manner. As of writing, it contains more than
100.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using  [Nix]{.pkg}, it is possible to install not only [R]{.proglang},
but also all the packages required for a project. The reason why one should use
[Nix]{.pkg} to install [R]{.proglang} packages, and not use the usual, built-in,
`install.packages()` [R]{.proglang} function instead, is that when installing a
package with  [Nix]{.pkg},  [Nix]{.pkg} makes sure to install every dependency of
every package, whether this dependency is another [R]{.proglang} package or a
system-level dependency.

@@

But why is installing software with  [Nix]{.pkg} reproducible? Because the
whole set of  [Nix]{.pkg} expressions is hosted on GitHhub, it is possible to use
a specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
specific commit of `nixpkgs` is called *pinning a revision (or commit)*. Pinning
a revision ensures that every package that Nix installs will always be at
exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`.

@@

With Nix, it is essentially possible to replace [renv]{.pkg} and
[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
`requirements.txt` files which list package dependencies for [Python]{.proglang}
projects and virtual environments. 
It is also possible to build multi-language environments, containing R
and Python, a LaTeX distribution, and even install one's favorite text editor to
edit the project's source code. Essentially, one can define a complete,
project-specific, and reproducible environment and use it interactively, or
non-interactively. It will always be possible to rebuild this environment in the
future (as long as the `nixpkgs` GitHub repository is online) to run an analysis
sometime in the future.

@@

## Reproducible development environments with Nix {#sec-repro-nix}

@@

```
let
  pkgs = import (fetchTarball 
    "https://github.com/NixOS/nixpkgs/archive/976fa336....tar.gz"
  ) {};
  system_packages = builtins.attrValues {
    inherit (pkgs) R ;
  };
in
  pkgs.mkShell {
    buildInputs = [ system_packages ];
    shellHook = "R --vanilla";
  }
```

@@

```
nix-shell -p R rPackages.rix"
```

to be dropped in a temporary [R]{.proglang} session where [rix]{.pkg} is
available. From there, it is possible to generate new  [Nix]{.pkg}
expressions to build new environments. For example, here is how one could use
[rix]{.pkg} to generate a `default.nix` file that installs version 4.3.1 of
[R]{.proglang} and the [dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't
show the generated expression here because of space):

@@

[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
system. This is quite useful if one wants to generate an expression and use that
expression in a CI/CD environment such as Github Actions. For example, the
repository that contains the source code to this
article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
compilation. Each time one of the authors pushed to the master branch, a Github
Actions runner would be started,  [Nix]{.pkg} would get installed on said
runner, and an environment containing the right packages was generated from a
`default.nix` file that is hosted on the same repository. The paper would then
get compiled using [Quarto]{.pkg} in this reproducible environment.

@@

Expressions generated with [rix]{.pkg} will point to a fork of `nixpkgs` instead
of the official, upstream repository.

@@

On versions of R released after version 3.6, these two lines return the
following vector: `r out_r`, but in versions prior to 3.6, the result will be
different. This is a documented change as shown in the R version 3.6 release
notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]

@@

It is possible to use `with_nix()` from an R session that was installed itself
via Nix, or from a regular R installation.

@@

There are many tools to make one’s analyses reproducible, but Nix distinguishes
itself from these in that its main concern is in deploying the complete set of
packages that are *closed under the ''depends on'' relation*: Nix does not
simply install one piece of software, but its dependencies, and their
dependencies. This makes Nix extremely useful for reproducible research, but
because it solves a complex problem, it is a complex tool. With [rix]{.pkg}, we
wish to make using Nix more accessible by providing a familiar interface and
user experience to [R]{.proglang} users. Because it is possible to build
reproducible development shells using  [Nix]{.pkg}, it can accomodate any
use case: running scripts, pipelines, interactive [shiny]{.pkg} applications,
and serve [plumber]{.pkg} application programming interfaces.

@@

Here is the Nix expression generated from the call to the `rix()`
function of section #sec-repro-nix:

>>>>

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of [R]{.proglang}needs to be handled
separately. [renv]{.pkg} will also not install system-level dependencies such
as [cmake]{.pkg} for [nloptr]{.pkg} on Linux or macOS, so these still need to
be handled separately, as described in [renv]{.pkg}’s Readme: 
\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

@@

The `.Rprofile` file gets read by [R]{.proglang}when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror. However, unless the user explicitly and
manually manages the different projects' libraries, using the Posit Package
Manager will install all the package for all the projects at the given date.

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.pkg} by the [R]{.proglang}infrastructure team [-@rlib2023]. [rig]{.pkg}
makes it easy to install different versions of [R]{.proglang}, so one could use
it to install the right version to reproduce an analysis, and then use one of
the listed packages above to install the right library of [R]{.proglang}
packages. However, this involves many manual steps and is thus error prone and
time-consuming.

@@

A common approach seems to be to work on a study interactively using a standard
installation of [R]{.proglang}and [R]{.proglang}packages, and after the study is done, take some time to
write the required `Dockerfile` to provide an easy way for future replicators to
reproduce the results of the study. But this approach does not solve the problem
of deploying the same environment on several machines, and so ideally authors
would develop using a containerized development environment. Another challenge
of using [Docker]{.pkg} is that, as stated already, [Docker]{.pkg} images are
actually almost complete Linux operating systems, so experience with Linux is
highly recommended if one wishes to properly write a `Dockerfile`. Finally, one
also needs to be careful and write the `Dockerfile` in such a way that the build
process will always result in the same image. This is only possible if a
specific digest of a base image is used instead of a tagged version, but in
practice, this is rarely done.

@@

## The [Nix]{.pkg} package manager {#sec-nix}

[Nix]{.pkg} is a package manager that can be used to install and build software
in a completely reproducible manner. As of writing, it contains more than
100.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using [Nix]{.pkg}, it is possible to install not only [R]{.proglang},
but also all the packages required for a project. [Nix]{.pkg} is actually the
package manager of a Linux distribution called NixOS, but can be installed as a
stand-alone tool for other Linux distributions, macOS, and on the Windows
Subsystem for Linux on Windows^[Which means that in practical terms, non-Nixos
Linux distributions and Windows can be considered as the same platform.].

The reason why one should use [Nix]{.pkg} to install [R]{.proglang} packages,
and not use the usual, built-in, `install.packages()` [R]{.proglang} function
instead, is that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

@@

But why is installing software with [Nix]{.pkg} reproducible? Because the whole
set of [Nix]{.pkg} expressions is hosted on GitHhub, it is possible to use a
specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
specific commit of `nixpkgs` is called *pinning a revision (or commit)*. Pinning
a revision ensures that every package that [Nix]{.pkg} installs will always be at
exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`.

@@

With Nix, it is essentially possible to replace [renv]{.pkg} and
[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
`requirements.txt` files which list package dependencies for [Python]{.proglang}
projects and virtual environments. It is also possible to build multi-language
environments, containing [R]{.proglang}and Python, a LaTeX distribution, and even install
one's favorite text editor to edit the project's source code. Essentially, one
can define a complete, project-specific, and reproducible environment and use it
interactively, or non-interactively. It will always be possible to rebuild this
environment in the future (as long as the `nixpkgs` GitHub repository is online)
to run an analysis sometime in the future.

@@

## Reproducible development environments with [Nix]{.pkg} {#sec-repro-nix}

@@

```
let
  pkgs = import (fetchTarball 
    "https://github.com/NixOS/nixpkgs/archive/976fa336....tar.gz"
  ) {};
  system_packages = builtins.attrValues {
    inherit (pkgs) [R]{.proglang};
  };
in
  pkgs.mkShell {
    buildInputs = [ system_packages ];
    shellHook = "R --vanilla";
  }
```

@@

```
nix-shell -p [R]{.proglang}rPackages.rix"
```

to be dropped in a temporary [R]{.proglang} session where [rix]{.pkg} is
available.^[`nix-shell -p` starts an interactive shell where the package
specified with the `-p` flag are present.] From there, it is possible to
generate new [Nix]{.pkg} expressions to build new environments. For example,
here is how one could use [rix]{.pkg} to generate a `default.nix` file that
installs version 4.3.1 of [R]{.proglang} and the [dplyr]{.pkg} and
[chronicler]{.pkg} packages (we don't show the generated expression here because
of space):

@@

[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
[rix]{.pkg} can generate [Nix]{.pkg} expression even if [Nix]{.pkg} is not installed on one's
system. This is quite useful if one wants to generate an expression and use that
expression in a CI/CD environment such as Github Actions. For example, the
repository that contains the source code to this
article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
compilation. Each time one of the authors pushed to the master branch, a Github
Actions runner would be started,  [Nix]{.pkg} would get installed on said
runner, and an environment containing the right packages was generated from a
`default.nix` file that is hosted on the same repository. The paper would then
get compiled using [Quarto]{.pkg} in this reproducible environment.

@@

As previously explained, when invoked, the [Nix]{.pkg} package manager will use
expressions hosted on the `nixpkgs` GitHub repository to build software.
However, when generating expressions with [rix]{.pkg} our fork
`rstats-on-nix/nixpkgs` is used instead.

Using a fork instead of the official upstream `nixpkgs` repository has several
advantages.

The first thing we need to stress, is that it is important to keep in mind that
the [Nix]{.pkg} package manager is first and foremost the package manager of a
Linux distribution called NixOS, and as such, several technical and governance
choice have to be made that limit what we, the authors of [rix]{.pkg} want to offer
to the users of `{rix}`. For our needs, having a fork gives us more flexibility.

For example, while it is technically possible for [Nix]{.pkg} to provide many versions
(or variants, in [Nix]{.pkg} jargon) of the same package (for example, you can
install the latest [Emacs]{.pkg} by installing the `emacs` package, or
[Emacs]{.pkg} 28 by installing `emacs28`) this ultimately depends on whether the
maintainer wishes to package several variants, or whether it is practical. In
the case of the [R]{.proglang} ecosystem, with more than 20'000 CRAN and
Bioconductor packages, that is not possible for the maintainers to provide
several variants of the same packages. So for a given `nixpkgs` commit, it is
not possible to easily install a specific version of an [R]{.proglang} package
that is not included in that particular `nixpkgs` commit. Instead, [rix]{.pkg}
can install it from source with this syntax:

```{r, eval = FALSE}
rix(..., r_pkgs = "dplyr@1.0.7", ...)
```

However, some packages might need specific fixes to be successfully installed
by [Nix]{.pkg} (as explained in the case of [xlsx]{.pkg} previously) and thus
trying to install packages like this might fail.

Furthermore, it isn't practical to update the whole of the [R]{.proglang}
packages set on [Nix]{.pkg} every day: so while CRAN and Bioconductor get
updates daily, the [R]{.proglang} packages set on [Nix]{.pkg} gets updated only
around new releases of R. Again, this is a consequence of [Nix]{.pkg} being
first and foremost the package manager of a Linux distribution with its own
governance; daily updates of more than 20'000 package is simply not feasible.

Because of these reasons, we decided to fork `nixpkgs`: because it is a fork, we
can afford to do things in a way that could not be possible or practical for
upstream.

The first thing this fork allows us to do is offer a daily snapshot of CRAN.
Every day, thanks to Github Actions, the [R]{.proglang} packages set gets
updated, and the result commited to a dated branch. This has been going on since
the 14th of December 2024 (see
[here](https://github.com/rstats-on-nix/nixpkgs/tree/2024-12-14)). So when users
set a date as in `rix(date = "2024-12-14", ...)` the specified dated branch from
the fork is going to get used.

But this doesn’t mean that we recommend you
use any date from the `rstats-on-nix/nixpkgs` fork: instead, each Monday,
another action uses this fork and tries to build a set of popular packages on
Linux and macOS, and only if this succeeds is the date added through a PR to the
list of available dates on `{rix}`!

The reason this is done like this is to manage another *risk* of the upstream
`nixpkgs`. As you know, `nixpkgs` is huge, and though the utmost care is taken
by contributors and the PR review process is very strict, it can happen that
updating packages breaks other packages. For example recently RStudio was in a
broken state due to an issue in one its dependencies, `boost`. This is not the
fault of anyone in particular: it's just that packages get updated and packages
that depend on them should get updated as well: but if that doesn't happen
quickly enough, the `nixpkgs` maintainer faces a conundrum. Either he or she
doesn't update the package because it breaks others, but not updating a package
could be a security vulnerability, or he or she updates the package, but now
others, perhaps less critical packages are broken and need to be fixed, either
by their upstream developers, or by the `nixpkgs` maintainer of said packages.
In the case of RStudio a fix was proposed and promptly merged, but if you wanted
to install RStudio during the time it took to fix it, you would have faced
an error message, which isn't great if all you want is use [Nix]{.pkg} shells as
development environments.

So for us, having a fork allows us to backport these fixes and so if you try to
install RStudio using the latest available date, which is `"2025-02-10"`, it’s
going to work, whereas if you tried to build it on that date using upstream
`nixpkgs` you'd be facing an error!

We spent quite some time backporting fixes: we went back all the way to
2019. The way this works, is that we start by checking out a `nixpkgs` commit on
selected dates, then we "update" the [R]{.proglang}packages set by using the Posit CRAN and
Bioconductor daily snapshots. Then, we backport as many fixes as possible, and
ensure that a selection of popular packages work on both x86-linux (which
includes Windows, through WSL) and aarch64-darwin (the M-series of Macs). Then
we commit everything to a dated branch of the `rstats-on-nix/nixpkgs` fork. You
can check out all the available dates by running: `rix::available_dates()`.
We're pretty confindent that you should not face any issues when using [Nix]{.pkg} to
build reproducible environments for R. However, should you face a problem, don't
hesitate to open an issue!

We have now packages and [R]{.proglang}versions working on Linux and macOS from March 2019
to now. See [this
repository](https://github.com/rstats-on-nix/daily_cran/blob/master/readme.md)
that contains the scripts that allowed us to do it. Backporting fixes was
especially important for Apple Silicon computers, as it took some time for this
platform to work correctly on Nix. By backporting fixes, we can now provide
olders versions of these packages for Apple Silicon as well!

Using this approach, our fork now contains many more versions of working R
packages than upstream. [rix]{.pkg} will thus likely keep pointing towards our fork
in the future, and not upstream anymore. This should provide a much better user
experience. An issue with our fork though, is that by backporting fixes, we
essentially create new [Nix]{.pkg} packages that are not included in upstream, and thus,
these are not built by Hydra, Nix's CI platform which builds binary packages. In
practice this means that anyone using our fork will have to compile many
packages from source. Now this is pretty bad, as building packages from source
takes quite some time. But fear not, because thanks to
[Cachix](https://www.cachix.org/) we now also have a dedicated binary cache of
packages that complements the default, public [Nix]{.pkg} cache! We provide instructions
on how to use Cachix, it's very easy, it's just running 2 additional commands
after installing Nix. Using Cachix speeds up the installation process of
packages tremendously. I want to give my heartfelt thanks to
[Domen Kožar](https://www.cachix.org/about) for sponsoring the cache!

Another thing we do with our fork is run an action every day at midnight, that
monitors the *health* of the [R]{.proglang}packages set. Of course, we don't build every
CRAN package, merely a handful, but these are among the most popular or the most
*at-risk* of being in a broken state. See
[here](https://github.com/rstats-on-nix/monitor_health/actions).

@@

On versions of [R]{.proglang}released after version 3.6, these two lines return the
following vector: `r out_r`, but in versions prior to 3.6, the result will be
different. This is a documented change as shown in the [R]{.proglang}version 3.6 release
notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]

@@

It is possible to use `with_nix()` from an [R]{.proglang}session that was installed itself
via Nix, or from a regular [R]{.proglang}installation.

@@

There are many tools to make one’s analyses reproducible, but [Nix]{.pkg} distinguishes
itself from these in that its main concern is in deploying the complete set of
packages that are *closed under the ''depends on'' relation*: [Nix]{.pkg} does not
simply install one piece of software, but its dependencies, and their
dependencies. This makes [Nix]{.pkg} extremely useful for reproducible research, but
because it solves a complex problem, it is a complex tool. With [rix]{.pkg}, we
wish to make using [Nix]{.pkg} more accessible by providing a familiar interface and
user experience to [R]{.proglang} users. Because it is possible to build
reproducible development shells using  [Nix]{.pkg}, it can accomodate any
use case: running scripts, pipelines, interactive [shiny]{.pkg} applications,
and serve [plumber]{.pkg} application programming interfaces.

@@

Here is the [Nix]{.pkg} expression generated from the call to the `rix()`
function of section #sec-repro-nix:
